### 2021-07-07 rad-icfp-2021/complexity.tex
.
fixes from Sam's comments section 6

\section{Complexity}
\label{sec:complexity}

@@

We follow a simple model similar to the one in \cite{griewank2008evaluating}.
We assume the cost is divided into $4$ elementary measures, being the number of MOVES, ADDS, MULTS, and NLOPS.
MOVES assumes a flat memory and represents moving a fixed size information  (e.g. 64 bits). 
ADDS represents the number of additions, 
MULTS the number of multiplications, 
and NLOPS the number of elementary non-linear operations like !cos! or !exp!.

@@

@@

As $\directD{\rho}{\Gamma}{Y}$ is defined by induction on programs, it suffices to show locally 
that $\directD{\rho}{\Gamma}{Y}$ verifies the cheap gradient principle.

@@

Combining the two previous points, we conclude that our reverse mode transformation satisfies a version of the cheap gradient principle.

\begin{theorem}
    Given a term !$\Gamma \vdash$ e: $\reals$! in Source with at most $p$ nested !map2! or !reduce!. 
    Denote by $G$ the term $\grad_\Gamma$!e! after inlining and forward substitution.
    Then !$\cost$(G) $\leq$ $4*2^{p}*\cost$(e)!.
\end{theorem}

\begin{proof}[notes]
    The proof is by routine induction on !e!. 
    First, one computes the cost of $\directD{\rho}{\Gamma}{Y}$(!e!).
    It has $N$ too many MOVES which are removed by inlining and forward substitution using the invariant that at each step, 
    at most 2 !x!$_i$ in the continuation are not variables.
\end{proof}

@@

In Figure~\ref{fig:optim} we present a of optimizations. 
As our language is purely functional and effect free, 
we can use these optimizations aggressively. 
A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
After these optimizations, the output program is essentially a sequence of let bindings 
and resembles SSA-code.

>>>>

\section{Complexity analysis}
\label{sec:complexity}

In this section, we introduce a simple cost model on our language and show 
that our reverse-mode transformation satisfies a version of the cheap gradient principle (Theorem \ref{thm:complexity}).

@@

We follow a simple model similar to the one in \cite{griewank2008evaluating}.
We assume the cost is divided into $4$ elementary measures, being the number of MOVES, ADDS, MULTS, and NLOPS.
MOVES assumes a flat memory and represents moving fixed-size information (e.g. 64 bits). 
ADDS represents the number of additions, 
MULTS the number of multiplications, 
and NLOPS the number of elementary non-linear operations like !cos! or !exp!.

@@

We define the nesting of array operations $\NestA$ of a term !e! of Source by induction as follows.

\begin{tabular}{r c l}
    $\NestA$(!c!) &=& 0 \\
    $\NestA$(!x!) &=& 0 \\
    $\NestA$(!let x:A = e$_1$ in e$_2$:B!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)  \\ 
    $\NestA$(!< e$_1$, e$_2$ >:AxB!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!) \\ 
    $\NestA$($\pi_i$(!e!)) &=& $\NestA$(!e!)\\
    $\NestA$(!e$_1$ op2 e$_2$!) &=& max $\NestA$(!e$_1$!), $\NestA$(!e$_2$!)\\
    $\NestA$(!op1 e!) &=& $\NestA$(!e!) \\
    $\NestA$(!map2 (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
    $\NestA$($\Gamma\vdash $ !reduce (x,y.e$_1$) e$_2$ e$_3$!) &=& max 1+$\NestA$(!e$_1$!),$\NestA$(!e$_2$!),$\NestA$(!e$_3$!) \\
\end{tabular}

\begin{theorem}
    \label{thm:complexity}
    Given a term !$\Gamma \vdash$ e: $\reals$! such that $\NestA$(!e!)$\leq$ $p$.
    Denote by $G$ the term $\grad_\Gamma$!e! after inlining and forward substitution.
    TODO: be more precise with "inlining and forward substitution".
    Then !$\cost$(G) $\leq$ $4*2^{p}*\cost$(e)!.
\end{theorem}

@@

\begin{proof}[Notes]
    As $\directD{\rho}{\Gamma}{Y}$ is defined by induction on programs, it suffices to show locally 
that $\directD{\rho}{\Gamma}{Y}$ verifies the cheap gradient principle.

@@

After that, the proof is by routine induction on !e!. 
First, one computes the cost of $\directD{\rho}{\Gamma}{Y}$(!e!).
It has $N$ too many MOVES which are removed by inlining and forward substitution using the invariant that at each step, 
at most 2 !x!$_i$ in the continuation are not variables.
\end{proof}

@@

In Figure~\ref{fig:optim} we present a of optimizations. 
These optimizations are all obviously valid according to the semantic model. 
As our language is purely functional, we can use these extra optimizations aggressively. 
A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
After these optimizations, the output program is essentially a sequence of let-binding
and resembles SSA-code.
